---
title: "LLM Practical Session: Clinical Applications"
author: "LV"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# Practical Workshop: LLMs in Clinical Applications

## 📋 Workshop Overview

This hands-on session will guide you through:

1. Setting up a local Large Language Model (LLM)
2. Testing clinical prompts with the model
3. Connecting to the model via API in R
4. Applying LLMs to clinical data extraction tasks

## 📥 Part 1: Setting up Local LLM (30 minutes)

### Setting up LM Studio

1. **Download and Install LM Studio**
   - Visit [https://lmstudio.ai/](https://lmstudio.ai/) and download the appropriate version
   - Follow installation instructions for your operating system
   
2. **Launch LM Studio and Download a Model**
   - Recommended models:
     - Mistral 7B Instruct v0.2
     - Llama 3 8B Instruct
     - Gemma 7B Instruct
   - These models balance capability and reasonable hardware requirements

3. **Start the Local Server**
   - In LM Studio, select your downloaded model
   - Click "Local Server" in the sidebar
   - Click "Start Server"
   - Note the URL (typically http://localhost:1234)

4. **Test the model**
   - In the Chat interface, ask: `"What is hepatitis?"`
   - Verify you get a reasonable medical response

## 🗣️ Part 2: Testing Chat Mode (40 minutes)

### Basic Clinical Prompts (5 minutes)

Use the LM Studio chat interface to test this prompt:

```
A patient presents with elevated AST and ALT. List possible causes.
```

### Multi-turn Clinical Conversation (10 minutes)

Create a dialogue:

1. Initial prompt:
```
A 45-year-old man presents with chest pain. What initial questions would you ask?
```

2. Follow up with:
```
The pain worsens with exertion, improves at rest.
```

### Chain-of-Thought Reasoning (10 minutes)

Test this prompt:

```
A patient has jaundice, right upper quadrant pain, and fever. Think step-by-step: what is the most likely diagnosis?
```

### Role Play Specialist Mode (10 minutes)

```
You are an infectious disease specialist. A 35-year-old man has persistent fever despite antibiotic therapy. Provide differential diagnoses considering his travel history to Africa.
```

### Critique Mode (5 minutes)

1. First prompt:
```
Summarize the key findings of this lab report: WBC 12,000/μL with left shift, hemoglobin 10.2 g/dL, platelets 450,000/μL, CRP 75 mg/L.
```

2. Then ask:
```
Now review your own summary. Identify possible mistakes or missing information.
```

### Ready-to-use Clinical Prompts Set

1. "Patient with fever, cough, and dyspnea: suggest differential diagnoses."
2. "Elevated D-dimer and chest pain: what would you recommend?"
3. "Interpret these liver function test results: AST 80, ALT 90, GGT 200."
4. "List common causes of microcytic anemia."
5. "Summarize the main treatments for community-acquired pneumonia."

### Discussion Points
- Note the model's reasoning capabilities
- Identify any hallucinations or inconsistencies
- Observe how specifying roles affects responses
- Evaluate the model's ability to self-critique

## 🧪 Part 3: Using the Local API in R (30 minutes)

### Setup: Connecting to LM Studio API

```{r}
# Install required packages
install.packages(c("httr", "jsonlite", "tibble", "dplyr", "stringr"))

library(httr)
library(jsonlite)
```

### Creating a Helper Function to Call the LM Studio API

```{r}
# Create a function to communicate with the LM Studio API
call_lm_studio <- function(prompt, 
                           system_prompt = NULL,
                           api_url = "http://localhost:1234/v1/chat/completions",
                           model = "local-model",
                           temperature = 0.7,
                           max_tokens = 1000) {
  
  # Create messages array
  messages <- list()
  
  # Add system message if provided
  if (!is.null(system_prompt)) {
    messages <- append(messages, list(list(
      role = "system",
      content = system_prompt
    )))
  }
  
  # Add user message
  messages <- append(messages, list(list(
    role = "user",
    content = prompt
  )))
  
  # Prepare request body
  body <- list(
    model = model,
    messages = messages,
    temperature = temperature,
    max_tokens = max_tokens
  )
  
  # Send POST request
  response <- httr::POST(
    url = api_url,
    body = jsonlite::toJSON(body, auto_unbox = TRUE),
    httr::add_headers("Content-Type" = "application/json"),
    encode = "json"
  )
  
  # Parse response
  parsed <- jsonlite::fromJSON(httr::content(response, "text", encoding = "UTF-8"))
  
  # Extract and return the content
  if (!is.null(parsed$choices) && length(parsed$choices) > 0) {
    return(parsed$choices[[1]]$message$content)
  } else {
    warning("No content returned from API")
    return(NULL)
  }
}
```

### Testing the API Connection

```{r}
# Test API connection with a simple clinical prompt
result <- call_lm_studio(
  prompt = "Summarize these lab results: Hb 10.5, MCV 70, RDW high.",
  system_prompt = "You are a helpful medical assistant."
)

# Print result
cat(result)
```

### Making Complex Medical Queries

```{r}
# Example of a more complex query
clinical_query <- "
A 67-year-old female presents with the following lab values:
- Hemoglobin: 9.8 g/dL (ref: 12.0-15.5)
- MCV: 76 fL (ref: 80-100)
- Ferritin: 15 ng/mL (ref: 30-400)
- TIBC: 450 μg/dL (ref: 250-370)

Provide an interpretation of these results and suggest possible diagnoses.
"

interpretation <- call_lm_studio(
  prompt = clinical_query,
  system_prompt = "You are a hematology specialist. Provide concise and evidence-based interpretations."
)

cat(interpretation)
```

## 📊 Part 4: Clinical-Laboratory Application Exercise (30 minutes)

### Mini-project: Clinical Data Extraction

In this exercise, we'll extract structured information from unstructured clinical trial descriptions.

```{r}
library(tibble)
library(dplyr)
library(stringr)

# Create a dataframe of clinical trial descriptions
trials_df <- tibble::tibble(
  trial_id = c("Trial 1", "Trial 2", "Trial 3"),
  text = c(
    "In a recent clinical trial, 200 patients were randomized to receive Drug A or placebo. The Drug A group had a 35% reduction in risk of hospitalization, whereas the placebo group showed no significant change. Adverse events occurred in 10% of the Drug A group and 5% of the placebo group.",
    
    "A double-blind study involving 150 patients compared Drug B against placebo. Patients on Drug B showed a 25% decrease in hospital admissions, while adverse events were reported in 12% of patients treated with Drug B versus 7% in the placebo group.",
    
    "Another randomized trial evaluated Drug C in 180 patients. Drug C recipients experienced a 40% lower hospitalization rate compared to placebo. However, adverse effects were slightly higher at 15% versus 8% in the placebo cohort."
  )
)

# Function to extract table from a single trial text
extract_table_from_trial <- function(trial_text) {
  # Create a prompt that specifies exactly what to extract
  prompt <- paste0(
    "Extract the information from the following study and format it into a markdown table.\n\n",
    trial_text,
    "\n\n",
    "Create columns: Treatment, Sample Size, Effect on Hospitalization (%), Adverse Events (%). Output only the markdown table without any additional text."
  )
  
  # Call the LLM
  result <- call_lm_studio(
    prompt = prompt,
    system_prompt = "You are a research assistant. Extract data precisely from clinical trials and format it as requested."
  )
  
  return(result)
}

# Initialize list to collect parsed tables
all_rows <- list()

# Iterate over each trial
for (i in seq_len(nrow(trials_df))) {
  cat(paste0("\nProcessing ", trials_df$trial_id[i], "...\n"))
  
  # Get the markdown table
  markdown_table <- extract_table_from_trial(trials_df$text[i])
  cat(markdown_table, "\n")
  
  # Parse markdown table into a dataframe
  # Match each row of the table
  rows <- str_match_all(markdown_table, "\\|\\s*(.*?)\\s*\\|\\s*(.*?)\\s*\\|\\s*(.*?)\\s*\\|\\s*(.*?)\\s*\\|")[[1]]
  
  if (nrow(rows) > 0) {
    # Skip header and separator rows (first two rows)
    data_rows <- rows[-(1:2), , drop = FALSE]
    
    if (nrow(data_rows) > 0) {
      # Create a temporary dataframe
      df_temp <- tibble::tibble(
        Trial = trials_df$trial_id[i],
        Treatment = data_rows[, 2],
        Sample_Size = data_rows[, 3],
        Effect_on_Hospitalization = data_rows[, 4],
        Adverse_Events = data_rows[, 5]
      )
      
      # Add to list
      all_rows[[length(all_rows) + 1]] <- df_temp
    }
  }
}

# Combine all tables
final_table <- bind_rows(all_rows)

# View combined final table
print(final_table)
```

### Additional Ideas for Clinical Applications

1. **Label Lab Results**
   - Extract abnormal results and tag them accordingly

2. **Summarize Patient History**
   - Create concise summaries from longer patient notes

3. **Generate Differential Diagnoses**
   - Based on symptoms, vital signs, and lab values

4. **Extract Medical Entities**
   - Identify diagnoses, treatments, medications from text

## 🔍 Discussion Points

1. **Model Limitations**
   - Identify when the model makes errors or hallucinates information
   - Determine what types of queries produce the most accurate responses

2. **Clinical Safety**
   - How would you validate the model's outputs?
   - What safeguards would you implement in a clinical setting?

3. **Comparison with Commercial Models**
   - How do these local models compare to commercial ones like GPT-4?
   - In what scenarios might a local model be preferable?

4. **Potential Applications**
   - Brainstorm useful applications in your specific clinical setting
   - What workflows could benefit most from LLM assistance?

## 📚 Resources

- [LM Studio Documentation](https://lmstudio.ai/docs)
- [Mistral AI Documentation](https://docs.mistral.ai)
- [OpenAI API Reference](https://platform.openai.com/docs/api-reference) (for API format compatibility)
- [httr Package Documentation](https://httr.r-lib.org)
- [jsonlite Package Documentation](https://cran.r-project.org/web/packages/jsonlite/index.html)