---
title: "LLM_practical_session"
author: "LV"
date: "`r Sys.Date()`"
output: html_document
---

# Practical Kit for "Generative AI and LLMs in Clinical-Laboratory Applications"

## 📅 2-hour Practice Session Content (English)

### Part 1: Setting up a Local LLM (30 minutes)
- **Tool**: LM Studio
- **Steps**:
  1. Install LM Studio.
  2. Download and launch a small open LLM model (e.g., *Mistral 7B*, *Llama 3 8B*).
  3. Confirm it is operational by asking: `"What is hepatitis?"`

---

### Part 2: Testing Chat Mode (40 minutes)

**Interact with the model via GUI.**

#### Basic Clinical Prompt (5 minutes)
- Prompt: "A patient presents with elevated AST and ALT. List possible causes."

#### Multi-turn Clinical Conversation (10 minutes)
- Prompt: "A 45-year-old man presents with chest pain. What initial questions would you ask?"
- Student inputs: "The pain worsens with exertion, improves at rest."
- Expected: Differential diagnosis suggestions.

#### Chain-of-Thought Reasoning (10 minutes)
- Prompt: "A patient has jaundice, right upper quadrant pain, and fever. Think step-by-step: what is the most likely diagnosis?"

#### Role Play Specialist Mode (10 minutes)
- Prompt: "You are an infectious disease specialist. A 35-year-old man has persistent fever despite antibiotic therapy. Provide differential diagnoses considering his travel history to Africa."

#### Critique Mode (5 minutes)
- Initial Prompt: "Summarize the key findings of this lab report."
- Follow-up Prompt: "Now review your own summary. Identify possible mistakes or missing information."

**Discussion Points:**
- Ability of the LLM to reason over time.
- Identification of hallucinations or inconsistencies.
- Importance of specifying roles and contexts.
- Observing LLMs' ability to self-critique.

**Ready-to-use Clinical Prompts Set:**
1. "Patient with fever, cough, and dyspnea: suggest differential diagnoses."
2. "Elevated D-dimer and chest pain: what would you recommend?"
3. "Interpret these liver function test results: AST 80, ALT 90, GGT 200."
4. "List common causes of microcytic anemia."
5. "Summarize the main treatments for community-acquired pneumonia."

---

### Part 3: Using the Local API with `ellmer` in R (30 minutes)

**Example R script to query the local LM Studio API:**

**First, install `ellmer` package:**

```{r}
install.packages("ellmer")

library(ellmer)

# Setup your local server URL (adjust port if needed)
llm <- ChatLLM(
  url = "http://localhost:1234/v1/chat/completions", # LM Studio default API endpoint
  model = "your-model-name-here" # e.g., "mistral-7b-instruct-v0.1"
)

# Create a prompt
response <- llm$complete(
  messages = list(
    list(role = "user", content = "Summarize these lab results: Hb 10.5, MCV 70, RDW high.")
  )
)

# View response
cat(response$content)
```


```{r connect-to-llm}
# Connect to the local LLM
llm <- ChatLLM(
  url = "http://localhost:1234/v1/chat/completions",
  model = "your-model-name-here"
)
```

```{r send-prompt}
# Send a clinical prompt
response <- llm$complete(
  messages = list(
    list(role = "user", content = "Summarize these lab results: Hb 10.5, MCV 70, RDW high.")
  )
)

# Print response
cat(response$content)
```

---

### Part 4: Clinical-Laboratory Application Exercise (30 minutes)

**Mini-project:**
- Form small groups.
- Each group receives 2 symptom sets or lab reports.
- Task: Write an optimal prompt and retrieve LLM suggestions.

**Discussion Points:**
- Identify hallucinations or unsafe recommendations.
- Discuss how to validate outputs against trusted clinical resources.

**Extended Clinical Data Extraction Task: Preparing a Meta-analysis**


```{r}
library(tibble)
library(ellmer)
library(dplyr)
library(stringr)

# Create a dataframe of studies
trials_df <- tibble::tibble(
  trial_id = c("Trial 1", "Trial 2", "Trial 3"),
  text = c(
    "In a recent clinical trial, 200 patients were randomized to receive Drug A or placebo. The Drug A group had a 35% reduction in risk of hospitalization, whereas the placebo group showed no significant change. Adverse events occurred in 10% of the Drug A group and 5% of the placebo group.",
    "A double-blind study involving 150 patients compared Drug B against placebo. Patients on Drug B showed a 25% decrease in hospital admissions, while adverse events were reported in 12% of patients treated with Drug B versus 7% in the placebo group.",
    "Another randomized trial evaluated Drug C in 180 patients. Drug C recipients experienced a 40% lower hospitalization rate compared to placebo. However, adverse effects were slightly higher at 15% versus 8% in the placebo cohort."
  )
)

# Connect to local LLM
llm <- ChatLLM(
  url = "http://localhost:1234/v1/chat/completions",
  model = "your-model-name-here"
)

# Function to extract table from a single trial text
extract_table_from_trial <- function(trial_text) {
  prompt <- paste0(
    "Extract the information from the following study and format it into a markdown table.\n\n",
    trial_text,
    "\n\n",
    "Create columns: Treatment, Effect on Hospitalization, Adverse Events (%)."
  )
  
  response <- llm$complete(
    messages = list(
      list(role = "user", content = prompt)
    )
  )
  
  return(response$content)
}

# Initialize list to collect parsed tables
all_rows <- list()

# Iterate over each trial
for (i in seq_len(nrow(trials_df))) {
  cat(paste0("\nProcessing ", trials_df$trial_id[i], "...\n"))
  
  markdown_table <- extract_table_from_trial(trials_df$text[i])
  
  # Parse markdown into a dataframe
  rows <- str_match_all(markdown_table, "\\|\\s*(.*?)\\s*\\|\\s*(.*?)\\s*\\|\\s*(.*?)\\s*\\|")[[1]]
  
  if (nrow(rows) > 0) {
    # Remove header line
    rows <- rows[-1, , drop = FALSE]
    
    df_temp <- tibble::tibble(
      Treatment = rows[, 2],
      Effect_on_Hospitalization = rows[, 3],
      Adverse_Events = rows[, 4]
    )
    
    all_rows[[trials_df$trial_id[i]]] <- df_temp
  }
}

# Combine all tables
final_table <- bind_rows(all_rows)

# View combined final table
final_table

```
